---
title: "Supporting Information:"
subtitle: "Biodiversity monitoring for a just planetary future"
format: pdf
highlight-style: pygments
#header-includes: \usepackage{setspace}\doublespacing
bibliography: references.bib
---

## Abstract:

While rapidly growing repositories of biodiversity data provide unprecedented insight into ecological patterns at global scales, the application of species observations often belies the reality that the species these data tell us most about is the one they were never intended to include: humans. Biodiversity data trace not only cities and roads but the rise of surveillance technology, shadows of colonial histories, and echoes of contemporary racial and economic disparities. However, these same data are increasingly used as the starting point to inform the implementation of global policy and the investment of billions of dollars to protect and restore nature over the next decade. Effectively leveraging large-scale biodiversity data to benefit both people and nature requires expertise in social, cultural, and political processes underlying data infrastructures and their histories, just as much as it requires more data and increasingly complex statistical methods.

In this SI document, we synthesize examples of the social, political and economic dimensions of human society reflected in global biodiversity data. We provide data, code, and citations for reproducing the figure in "Biodiversity monitoring for a just planetary future" (doi:10.1126/science. adh8874)

## Exploring social and political dimensions of biodiversity data

In this figure, we leverage the Global Biodiversity Information Facility (GBIF) occurrence data set to reproduce, visualize, update, and/or expand upon the cited social and political dimensions of biodiversity data presented in the paper.

GBIF releases full occurrence "snapshots" monthly. In this paper, we leverage the Sept 31, 2023 Snapshot, which was the most recent at the time of submission (<https://doi.org/10.15468/dl.ua9nww>) [@gbif.orguser2023]. This snapshot has approximate 2.6 billion occurrence records.

We provide code to reproduce each panel of Figure 1. All data used is accessed within the code and is openly available.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE, # show warnings
  message = FALSE, # show messages
  error = TRUE, # do not interrupt generation in case of errors,
  echo = TRUE  # show R code
)
```

We use the following packages:

```{r}
library(geomtextpath)
library(duckdbfs)
library(gbifdb) 
library(tidyverse) 
library(fst) 
library(sf) 
library(terra)
#library(raster)
library(MetBrewer)
library(rnaturalearth)
library(countrycode)
library(arrow)
library(usmap)
library(svglite)
```

Package citations: [@geomtextpath; @duckdb; @duckdbfs; @gbifdb; @tidyverse; @sf; @terra; @raster; @MetBrewer; @rnaturalearth; @countrycode; @arrow; @usmap]:

## Connect to a GBIF snapshot

We use a local copy of the Sept 31, 2023 GBIF snapshot and the `gbifdb` package [@gbifdb] to query the \>2.6 billion observations in the database.

```{r snapshot}
gbif <- gbif_local("/home/shared-data/gbif/occurrence/2023-10-01/occurrence.parquet/",
                   backend="duckdb")
```

All analysis here can alternatively be done by querying the GBIF [AWS snapshot](https://registry.opendata.aws/gbif/) (leveraging the `arrow` package [@arrow]) using the following code:

```{r}
 #snapshot <- "s3://gbif-open-data-eu-central-1/occurrence/2023-10-01/occurrence.parquet"
 #gbif <- open_dataset(gbif_snapshot)
```

## Panel A: Global map

We summarize the count of observations at 0.1 decimal degrees. All observations in GBIF with coordinates are included in this map.

```{r global_query, cache = TRUE}
df <- gbif |> 
  mutate(latitude = round(decimallatitude,2), 
         longitude = round(decimallongitude,2)) |> 
  #filter(year >1800) |>
  count(longitude, latitude) |> 
  collect()
```

We convert the lat/long to spatial points using the `sf` package [@sf].

```{r global_sf, cache = TRUE}
df_spatial <- df |> 
  filter(!is.na(latitude), 
         !is.na(longitude)) |> 
  st_as_sf(coords = c("longitude", "latitude"), 
           crs = "+proj=robin +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m")
```

The log of the sum of observations at each point is converted into a global raster at 0.1 degrees.

```{r global_raster, cache = TRUE}
library(raster)
ras_temp <-raster(xmn=-180, xmx=180, ymn=-90, ymx=90,
                  resolution=c(0.1,0.1), vals=NA)
global_plot_all <- rasterize(df_spatial, ras_temp, 
                         field = "n", fun='sum') 
#rm(df_spatial) #remove unnecessary data
rm(ras_temp) #remove unnecessary data
```

Reproject to the Robinson projection and plot using the terra package [@terra].

```{r global_plot, cache=TRUE}
crs <- "+proj=robin +lon_0=0 +x_0=0 +y_0=0 +datum=WGS84 +units=m" 
global_plot_all <- terra::rast(global_plot_all) 
global_plot <- global_plot_all * 1 # to deal with NAs in this dataset 
# reproject for viz
global_plot_r <- terra::project(global_plot, crs, mask=TRUE) 
# define color gradient
colors <- c("grey", met.brewer(name="Isfahan1",n=20,type="continuous"))
# take log for viz
terra::plot(log(global_plot_r,10), col = colors, axes = FALSE)
writeRaster(global_plot_r, "../data/panels/PanelA_data.tif", overwrite=TRUE)
```

```{r global_plot_svg, eval = FALSE, include = FALSE}
svg("../data/panels/panelA.svg")
terra::plot(log(global_plot_r,10), col = colors, axes = FALSE)
dev.off()
```

## Panel B: Macroeconomic patterns

In panel B, we show the cumulative number of observations per hectare collected in countries across different income groups. The World Bank classifies economies for analytical purposes into four income groups: **low, lower-middle, upper-middle, and high income**. We use these for our analysis.

```{r country_income, cache = TRUE}
world <- ne_countries(type = "countries", scale = "medium")
world <- st_as_sf(world) |>
  dplyr::select(iso_a2, income_grp) |>
  st_make_valid() |>
  mutate(area = st_area(geometry)) |>
  as_tibble() |>
  dplyr::select(-geometry) |>
  mutate(area = as.numeric(area)/10000) |>
  rename(countrycode = iso_a2)
```

Collect count of observations per year per country

```{r query_country_yr, cache = TRUE}
country_year <-  gbif |>
    count(countrycode, year) |>
    collect()
```

```{r macroecon_density}
macroeconomics_density_cumulative <- country_year |>
  filter(year >1949 & year < 2022) |>
  mutate(n = replace_na(n, 0)) |>
  left_join(world) |>
  mutate(income_grp = str_sub(income_grp, 4, -1),
         #INCOME_GRP = gsub("\\s", "\n", INCOME_GRP),
         income_grp = gsub("\\:.*","",income_grp)) |>
  group_by(year, income_grp) |> 
  summarise(n = sum(n, na.rm = TRUE), 
            area = sum(area)) |> ungroup()|>
  group_by(income_grp) |>
  mutate(cum_obs = cumsum(n)) |>
  ungroup() |>
  mutate(density = cum_obs/area) |>
  drop_na() |>
  ggplot(aes(year, density, color = income_grp, label = income_grp)) + 
  geom_line() +
  #geom_textline(size = 3, fontface = 2, spacing = 30, text_smoothing = 50) + 
  theme_classic() +
  theme(legend.position = "bottom", legend.title = element_blank()) +
  labs(x = "", y = "observations per ha (cumulative)") +  
  scale_color_manual(values=met.brewer("Homer2", 4)) +
  theme(legend.background = 
          element_rect(colour = 'black', fill = 'white', linetype='solid'))
macroeconomics_density_cumulative
```

```{r include=FALSE}
# for editors
svg("../data/panels/panelB.svg", height=3, width=3)
macroeconomics_density_cumulative
dev.off()
```

```{r include=FALSE}
# for editors
country_year |>
  filter(year >1949 & year < 2022) |>
  mutate(n = replace_na(n, 0)) |>
  left_join(world) |>
  mutate(income_grp = str_sub(income_grp, 4, -1),
         #INCOME_GRP = gsub("\\s", "\n", INCOME_GRP),
         income_grp = gsub("\\:.*","",income_grp)) |>
  group_by(year, income_grp) |> 
  summarise(n = sum(n, na.rm = TRUE), 
            area = sum(area)) |> ungroup()|>
  group_by(income_grp) |>
  mutate(cum_obs = cumsum(n)) |>
  ungroup() |>
  mutate(density = cum_obs/area) |>
  drop_na() |>
  write_csv("../data/panels/panelB_data.csv")
```

## Panel C: Redlining

### Redlining data

Redlining data is downloaded from Mapping Inequality (<https://dsl.richmond.edu/panorama/redlining/>) [@holc]. Here we reproduce the patterns found in Ellis-Soto et al., 2023 [@ellis-soto2023] (with an updated snapshot of GBIF data) showing disparities in the density of bird data throughout redlined neighborhoods in the United States.

In the figure, we compare neightborhoods that were deemed (holc grade A) to those deemed most hazardous (holc grade D).

```{r cache = TRUE}
# download redlining geojson 
holc <- 
  st_read("https://dsl.richmond.edu/panorama/redlining/static/mappinginequality.json") |>
  rename(holc_grade = grade) |>
  dplyr::select(state, city, holc_grade, geometry) |>
  dplyr::filter(!is.na(holc_grade) & holc_grade != 'E') |>
  sf::st_make_valid() |>
  mutate(holc_grade = ifelse(holc_grade == "A ", "A",
                             ifelse(holc_grade == "B ", "B",
                                    ifelse(holc_grade == "C ", "C",
                                           ifelse(holc_grade == "D ", "D", holc_grade))))) 
# remove inalid polygons 
holc <- holc |>
  dplyr::mutate(valid =st_is_valid(holc)) |> 
  dplyr::filter(valid=="TRUE")
# calculate area per neighborhood
holc_area  <- holc |>
  mutate(area = st_area(geometry)) |>
  as_tibble() |> drop_na() |>
  dplyr::select(-geometry) |>
  group_by(holc_grade) |>
  summarise(area = sum(area)) |>
  group_by(holc_grade) |>
  summarise(area = sum(area)) |>
  mutate(area = as.numeric(area)/10000) # area in ha
```

### US observations

Species observations in class "Aves" in the US are queried and summerized at 0.001 decimal degrees.

```{r cache = TRUE}
states <- sf::st_as_sf(maps::map("state", 
                                 plot = FALSE, fill = TRUE)) 
ext_states <- ext(states) 

US_pts <- gbif |>
   filter(countrycode == "US") |>
   filter(class == "Aves") |>
   mutate(latitude = round(decimallatitude,3), 
         longitude = round(decimallongitude,3)) |>
  count(latitude, longitude) |>
  collect() |>
  filter(longitude > ext_states[1] & longitude < ext_states[2] &
           latitude > ext_states[3] & latitude < ext_states[4]) 
```

These points are converted to a spatial data frame and reprojected. For the US map we plot a subset of 1 million points to allow for visualization of point density in cities and around major roads.

```{r cache = TRUE}
US_pts_sf_all <- US_pts |>
  filter(!is.na(latitude), 
         !is.na(longitude)) |>
  st_as_sf(coords = c("longitude", "latitude"), 
           crs = st_crs(holc)) 

US_pts_sf <- US_pts_sf_all |>  head(1000000) |>
  st_transform(crs = usmap::usmap_crs())
```

```{r cache = TRUE, fig.width=10, fig.height=10}
plot_gbif <- US_pts_sf |>
  ggplot() +
  geom_sf(aes(geometry = geometry), alpha = 0.05,
            size = 0.001, color= "black") +
  theme(legend.position = "none") +
  theme_void()

plot_gbif
```

```{r include=FALSE}
# for editors
svg("../data/panels/panelC_US.svg", height=10, width=10)
plot_gbif
dev.off()
```

### Redlining observations summary

The bar chart in Panel C shows the number of aves observations per unit area in grade A vs holc grade D across all cities included in the Mapping Inequality dataset.

```{r cache = TRUE}
holc_obs <- st_join(US_pts_sf_all, holc, join = st_within)
```

```{r}
redline_area <- holc_area |>
  filter(holc_grade %in% c("A", "D")) |>
  #filter(holc_grade %in% c("A", "B", "C", "D")) |>
  mutate(redlined = ifelse(holc_grade == "D", "redlined", "not redlined")) |>
  group_by(redlined) |>
  summarise(area = sum(area))
  
redlining <- as_tibble(holc_obs) |>
  drop_na() |>  
  filter(holc_grade %in% c("A", "D")) |>
  #filter(holc_grade %in% c("A", "B", "C", "D")) |>
  mutate(redlined = ifelse(holc_grade == "D", "redlined", "not redlined")) |>
  group_by(redlined) |>
  summarise(counts = sum(n)) %>% ungroup() |>
  left_join(redline_area) |>
  mutate(density = counts/area) |> drop_na() |> 
  ggplot(aes(x = redlined, y = density, fill = redlined)) + 
  scale_fill_manual(
    values=c("grey", "firebrick4")) +
  geom_col(width = 0.6, alpha = 0.7) + theme_classic() + 
  theme(legend.position = "none") + 
  labs(y = " sampling density \n (obs per hectare;  birds)", x = "")

as_tibble(holc_obs) |>
  drop_na() |>  
  filter(holc_grade %in% c("A", "D")) |>
  #filter(holc_grade %in% c("A", "B", "C", "D")) |>
  mutate(redlined = ifelse(holc_grade == "D", "redlined", "not redlined")) |>
  group_by(redlined) |>
  summarise(counts = sum(n)) %>% ungroup() |>
  left_join(redline_area) |>
  mutate(density = counts/area) |> drop_na() |>
  write_csv("../data/panels/redlining_simplified.csv")

redlining
```

```{r include=FALSE}
# for editors
svg("../data/panels/redlining.svg", height = 3, width = 3)
redlining
dev.off()
```

```{r include = FALSE}
as_tibble(holc_obs) |>
  drop_na() |>  
  group_by(holc_grade) |>
  summarise(counts = sum(n)) |>
  ungroup() |>
  left_join(holc_area) |>
  mutate(density = counts/area) |>
  dplyr::select(holc_grade, density) |>
  write_csv("../data/panels/panelC_holc_bar_data.csv")
```

## Panel D: Conflict

Following analysis in [@zizka2021] and leveraging the yearly conflict data from the Uppsala Conflict Data Program (UCDP) [@davies2023; @gleditsch2002], we show how biodiversity data observations track conflict both (i) globally and (ii) in Cambodia [@zizka2021] since 1950.

```{r cache = TRUE}
country_year <-  gbif |>
    count(countrycode, year) |>
    collect()
```

```{r}
#download.file(url = "https://ucdp.uu.se/downloads/ucdpprio/ucdp-prio-acd-231-rds.zip", destfile = "data/conflict.zip")
#unzip("data/conflict.zip", exdir = "data/")

conflict <- readRDS("../data/UcdpPrioConflict_v23_1.rds") |>
  dplyr::select(location, year, intensity_level) |>
  separate_rows(location, sep = ",") |>
  mutate(countrycode = 
           countrycode(location, 
                       origin = 'country.name', 
                       destination = 'iso2c'))
```

```{r}
conflict_plot_i <- country_year |>
  left_join(conflict) |>
  dplyr::select(-location) |>
  filter(year >1950 & year <2022) |>
  mutate(intensity_level = replace_na(intensity_level, 0)) |>
  mutate(n = replace_na(n, 0)) |>
  group_by(year, intensity_level) |>
  summarise( n = sum(n, na.rm = TRUE)) |>
  unique() |>
  ggplot() + 
  geom_line(aes(year, log(n,10), col = as.factor(intensity_level), group=as.factor(intensity_level))) + 
  #scale_color_manual(values= c("darkgrey", "#FF4433", "darkred")) +
  scale_color_manual(values=met.brewer("Renoir", 4)) +
  #geom_line(aes(year, v2xcl_dmove*10), lwd = 1.5, color = "black") + 
  theme_classic() + 
  theme(legend.position = "none", axis.title.x = element_blank()) +
  scale_y_continuous(
    # Features of the first axis
    name = "log(# observations)"#,
    # Add a second axis and specify its features
    #sec.axis = sec_axis( trans=~./10, name="freedom of movement")
  ) #+ geom_line(aes(x = year, y = intensity_level*5)) +

conflict_plot_i
```

```{r}
conflict_plot_ii <- country_year |>
  filter(countrycode == "KH") |> filter(year >1949 & year < 2022) |> arrange(-year) |>
  left_join(conflict, by = c("countrycode", "year")) |>
  dplyr::select(-location) |>
  filter(year >1949 & year < 2022) |>
  mutate(intensity_level = replace_na(intensity_level, 0)) |>
  mutate(n = replace_na(n, 0)) |>
  group_by(countrycode, year) |> count()
  summarise(intensity_level = max(intensity_level),
            n = mean(n, na.rm = TRUE)) |>
  unique() |>
  ggplot() + 
  geom_col(aes(year, (log(n)), fill = as.factor(intensity_level))) + 
  #scale_fill_manual(values= c("darkgrey", "#FF4433", "darkred")) +
  scale_fill_manual(values=met.brewer("Renoir", 4)) +
  #geom_line(aes(year, v2xcl_dmove*10), lwd = 1.5, color = "black") + 
  theme_classic() + 
  theme(legend.position = "none", axis.title.x = element_blank()) +
  scale_y_continuous(
    # Features of the first axis
    name = "log(# observations)"#,
    # Add a second axis and specify its features
    #sec.axis = sec_axis( trans=~./10, name="freedom of movement")
  ) #+ geom_line(aes(x = year, y = intensity_level*5)) +

conflict_plot_ii
```

```{r include = FALSE}
country_year |>
  left_join(conflict) |>
  dplyr::select(-location) |>
  filter(year >1950) |>
  mutate(intensity_level = replace_na(intensity_level, 0)) |>
  mutate(n = replace_na(n, 0)) |>
  filter(countrycode == "KH") |>
  group_by(countrycode, year) |>
  summarise(intensity_level = max(intensity_level),
            n = mean(n, na.rm = TRUE)) |>
  write_csv("../data/panels/panelD_ii_data.csv")
```

```{r}
country_year |>
  left_join(conflict) |>
  dplyr::select(-location) |>
  filter(year >1949 & year <2022) |>
  mutate(intensity_level = replace_na(intensity_level, 0)) |>
  mutate(n = replace_na(n, 0)) |>
  group_by(year, intensity_level) |>
  summarise(n = sum(n, na.rm = TRUE)) |>
  unique() |>
  write_csv("../data/panels/panelD_i_data.csv")
```

```{r include = FALSE}
svg("../data/panels/panelD_i.svg", height=2.5, width=3)
conflict_plot_i
dev.off()
```

```{r include = FALSE}
svg("../data/panels/panelD_ii.svg", height=2.5, width=3)
conflict_plot_ii
dev.off()
```

## Panel E: Colonialism

As explored in [@zizka2021], social and political factors impact who has collected biodiversity data. We reproduce this analysis using the update GBIF snapshot. We can see that the publishing country before and after Nigeria's independence (1960) is drastically different [@zizka2021].

```{r cache = TRUE}
NG_year <- gbif |>
    filter(countrycode == "NG") |>
    count(year, basisofrecord, datasetkey) |>
    collect()

# download dataset keys to keep track of publishing country
orgs <- read_tsv("https://api.gbif.org/v1/dataset/search/export?format=TSV&") |>
  dplyr::select(publishing_country, dataset_key, title) |>
  rename(datasetkey = dataset_key)

NG_year_summary <- NG_year |>
  left_join(orgs) |>
  mutate() |>
  mutate(country_data = ifelse(publishing_country == "GB", "GB", 
                               ifelse(publishing_country == "NG", "NG", "other"))) |>
  mutate(precol = ifelse(year <1961,"1","2")) |>
  group_by(country_data, precol) |>
  summarise(n = sum(n)) |>
  drop_na()
```

### Observations in Nigeria pre-independence

```{r}
t1 <- NG_year_summary |> 
  group_by(precol) |> mutate(total_obs = sum(n)) |> ungroup()|>
  mutate(perc_obs = n/total_obs) |> filter(precol == "1") |>
  arrange(desc(perc_obs)) %>%
  mutate(lab.pos = cumsum(perc_obs)-.5*perc_obs)

panelE_i <- ggplot(data = t1, 
       aes(x = 2, y = perc_obs, fill = country_data))+
  geom_bar(stat = "identity")+
  coord_polar("y", start = 200) +
  theme_void() +
  scale_fill_brewer(palette = "Dark2") + xlim(.2,2.5)

panelE_i
```

```{r include = FALSE}
write_csv(t1, "../data/panels/panelE_i.csv")

svg("../data/panels/panelE_i.svg", height=3, width=5)
panelE_i
dev.off()
```

### Observations in Nigeria post-independence

```{r}
t2 <- NG_year_summary |> 
  group_by(precol) |> mutate(total_obs = sum(n)) |> ungroup()|>
  mutate(perc_obs = n/total_obs) |> filter(precol == "2") |>
  arrange(desc(perc_obs)) %>%
  mutate(lab.pos = cumsum(perc_obs)-.5*perc_obs)

panelE_ii <- ggplot(data = t2, 
       aes(x = 2, y = perc_obs, fill = country_data))+
  geom_bar(stat = "identity")+
  coord_polar("y", start = 200) +
  theme_void() +
  scale_fill_brewer(palette = "Dark2") + xlim(.2,2.5)

panelE_ii
```

```{r include = FALSE}
write_csv(t2, "../data/panels/panelE_ii.csv")

svg("../data/panels/panelE_ii.svg", height=3, width=5) 
panelE_ii 
dev.off()
```

## References
